---
title: "Bayesian RMST Network Meta-Analysis Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian RMST Network Meta-Analysis Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This tutorial walks through a complete analysis using the `rmstnma` package, 
from data preparation to results visualization for Bayesian RMST network 
meta-analysis.

## Prerequisites

You need `rmstnma` and a working installation of **CmdStan**.

```r
library(rmstnma)
library(ggplot2)
```

## 1. Prepare the Network

Start by loading your survival data. We'll use the package's `example_network`.

```{r}
data(example_network)
head(example_network)

# Create a network object
net <- rmst_network(
  data = example_network,
  study = "study_id",
  trt = "treatment",
  time = "time",
  surv = "survival"
)

# Visualize the network structure
# plot(net, type = "network")
```

## 2. Fit the Model

We'll fit a Bayesian model using `rmst_nma()`. We specify `tau = 24` months 
as our restriction time.

```{r, eval=FALSE}
# Note: Requires CmdStan
fit <- rmst_nma(
  network = net,
  tau = 24,
  baseline = "rp_spline",
  chains = 4,
  iter = 2000
)
```

## 3. Examine Results

Once fitted, you can generate various plots to interpret the treatment effects.

### Forest Plot
Compare treatments against a reference (defaults to the first treatment in the network).

```{r, eval=FALSE}
plot(fit, type = "forest", reference = "Control")
```

### Rankogram and SUCRA
Estimate the probability that each treatment is ranked 1st, 2nd, etc.

```{r, eval=FALSE}
plot(fit, type = "rank")
```

### Sensitivity to tau
RMST depends on the restriction time. It's good practice to fit the model 
across multiple tau values.

```{r, eval=FALSE}
fit_multi <- rmst_nma(network = net, tau = c(12, 18, 24, 30, 36))
plot(fit_multi, type = "rmst")
```

## 4. Convergence Diagnostics

Always check your MCMC chains for convergence.

```{r, eval=FALSE}
plot(fit, type = "trace", pars = c("d[2]", "sigma"))
```

## Conclusion

The `rmstnma` package simplifies the complex process of Bayesian survival NMA, 
providing a robust and flexible framework for comparative effectiveness research.
